Dependent Object Types

### 2012-09-03 doc/candidacy/report.tex
feedback of eugene

We present three papers~\cite{homl,tribe,nuObj}, each of which models
a calculus with path-dependent types.
\begin{itemize}
\item The first paper, {\it A Type-Theoretic Approach to Higher-Order
  Modules with Sharing} by Robert Harper and Mark
  Lillibridge~\cite{homl}, presents a calculus for modules in the ML
  tradition. Path-dependent types arise because modules, which may
  contain type components, are first-class values in their calculus.
\item The second paper, {\it Tribe: a simple virtual class calculus}
  by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias
  Wrigstad~\cite{tribe}, presents a core calculus for virtual
  classes. Path-dependent types arise because, in Tribe, paths are
  types, and, almost vice versa: each type is a generalized form of
  path.
\item The third paper, {\it A Nominal Theory of Objects with Dependent
  Types} by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
  Matthias Zenger~\cite{nuObj}, presents a calculus which unifies
  concepts from advanced object-oriented systems and module systems
  and, with~\cite{FS}, serves as a type-theoretic foundation for many
  features of Scala's type system. Path-dependent types arise because
  any path $p$ has a singleton type $p.\type$.
\end{itemize}

@@

In~\cite{homl}, Harper and Lillibridge present a calculus for
first-class modules, with complete control over the propagation of
compile-time information.

@@

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an if statement. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness, which
we illustrate with an example below. Another limitation of the SML
module system is that ``type sharing'' is limited to equality between
type names, not general type expressions. In effect, this means that
the transparency / opaqueness of a signature cannot be fully
fine-tuned. Their proposal introduces translucent signatures, which
allows any type to be partially or fully determined by a type
expression.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and ``out'' selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
subclassing relationship with one another, as this would only bring
confusion when ascribing a coloured graph to a graph. Emulating
``out'' faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the paper, using ``out'' in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

The terms in ${\nu}Obj$ denotes objects or classes. They consist of
variables $x$ denoting objects, selections $t.l$, object creations
\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
and mixin compositions $t \withsub{S} u$ denoting a class of self
type $S$ combined from two classes $t$ and $u$. A value is a variable
or a class template. A path is a variable followed by a possibly empty
sequence of selections.

The types in ${\nu}Obj$ consists of singleton types $p.\type$, type
selections $T\Tsel L$, record types $\obj{x \sep \tup D}$, compound
types $T \with U$, and class types $\cls{x: S \sep \tup D}$.

@@

Our immediate goal is to prove type soundness of the DOT calculus. We
first tried to prove the calculus type-safe using the standard
theorems of preservation and progress~\cite{soundness,
  tapl}. Unfortunately, for the DOT calculus and any variants that we
devise, preservation doesn't hold. Indeed, we found many
counterexamples to preservation due to narrowing, i.e.\ a type becoming more
precise after substitution. However, the standard theorems of
preservation and progress are just one way to prove type soundness. We
are exploring a promising avenue to prove type safey using the
powerful method of step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
proving the calculus sound, we are also refining it.

>>>>

We present three papers~\cite{homl,tribe,nuObj}, each of which models
a calculus with path-dependent types.
\begin{itemize}
\item The first paper, {\it A Type-Theoretic Approach to Higher-Order
  Modules with Sharing} by Robert Harper and Mark
  Lillibridge~\cite{homl}, presents a calculus for modules in the ML
  tradition. Path-dependent types arise because modules, which may
  contain type components, are first-class values in their calculus.
\item The second paper, {\it Tribe: a simple virtual class calculus}
  by Dave Clarke, Sophia Drossopoulou, James Noble and Tobias
  Wrigstad~\cite{tribe}, presents a core calculus for virtual
  classes. Path-dependent types arise because, in Tribe, paths are
  types, and, almost vice versa: each type is a generalized form of
  path.
\item The third paper, {\it A Nominal Theory of Objects with Dependent
  Types} by Martin Odersky, Vincent Cremet, Christine R{\"o}ckl and
  Matthias Zenger~\cite{nuObj}, presents a calculus which unifies
  concepts from advanced object-oriented systems and module systems
  and, with~\cite{FS}, serves as a type-theoretic foundation for many
  features of Scala's type system. Path-dependent types arise because
  any path $p$ has a singleton type $p.\type$, from which type members
  can be selected.
\end{itemize}

@@

In~\cite{homl}, Harper and Lillibridge present a calculus for
first-class modules, with complete control over the propagation of
information between compile-time program units.

@@

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an if expression. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness, which
we illustrate with an example below. Another limitation of the SML
module system is that ``type sharing'' is limited to equality between
type names, not general type expressions. In effect, this means that
the transparency / opaqueness of a signature cannot be fully
fine-tuned. Their proposal introduces translucent signatures, which
allows any type to be partially or fully determined by a type
expression.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and ``out'' selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
subclassing / further-binding with one another, as this would only bring
confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
``out'' faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the Tribe paper, using ``out'' in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

The terms in ${\nu}Obj$ denote objects or classes. They consist of
variables $x$ denoting objects, selections $t.l$, object creations
\mbox{$\newin{x \larrowc t}{u}$} defining an object $x$ of class $t$
whose scope is the term $u$, class templates $\cls{x: S \sep \tup d}$,
and mixin compositions $t \withsub{S} u$ denoting a class of self
type $S$ combined from two classes $t$ and $u$. A value is a variable
or a class template. A path is a variable followed by a possibly empty
sequence of selections.

The types in ${\nu}Obj$ consist of singleton types $p.\type$, type
selections $T\Tsel L$, record types $\obj{x \sep \tup D}$, compound
types $T \with U$, and class types $\cls{x: S \sep \tup D}$.

@@

Our immediate goal is to prove type soundness of the DOT calculus. We
first tried to prove the calculus type-safe using the standard
theorems of preservation and progress~\cite{soundness,
  tapl}. Unfortunately, for the DOT calculus and any variants that we
devise, preservation doesn't hold. Indeed, we found many
counterexamples to preservation due to narrowing, i.e.\ a type becoming more
precise after substitution. However, the standard theorems of
preservation and progress are just one way to prove type soundness. We
are exploring a promising avenue to prove type safety using the
powerful method of step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In the process of
proving the calculus sound, we are also refining it.
